---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

In this section I unzip the data from the repository, load the data into R and transform the date variable to date format.

```{r lib, echo=FALSE, include=FALSE}

library(ggplot2)
library(tidyverse)
library(lubridate)

```

```{r load, echo=TRUE}

unzip("activity.zip")
dat <- read.csv("activity.csv")

dat <- dat %>%
  mutate(date=as.Date(date))

```


## What is mean total number of steps taken per day?


```{r histogram, echo=TRUE}

dat_day <- dat %>%
  filter(!is.na(steps)) %>%
  group_by(date) %>%
  summarise(steps_day=sum(steps))

g <- ggplot(dat_day, aes(x=date, y=steps_day))
g+geom_col()+theme_bw()+labs(title="Total number of steps per day")


```
```{r mean, echo=TRUE}

summary(dat_day)

mean <- mean(dat_day$steps_day)
median <- median(dat_day$steps_day)

```

* The mean total number of steps taken per day is `r mean`
* The median total number of steps taken per day is `r median`

## What is the average daily activity pattern?

```{r daily_average, echo=TRUE}

dat_int <- dat %>%
  filter(!is.na(steps)) %>%
  group_by(interval) %>%
  summarise(steps_int=mean(steps),
            steps_med=median(steps))

plot(dat_int$interval, dat_int$steps_int, type="l", xlab="5-min interval", ylab="Average number of steps", main="Average daily activity pattern")

dat_int$interval[dat_int$interval==max(dat_int$steps_int)]

```

## Imputing missing values

### Total number of missing values

```{r na, echo=TRUE}

nas <- nrow(dat[is.na(dat$steps),])

```
Total number of rows with NAs is `r nas`.

### Fill in missing values

I use the median for the 5 minute interval to fill in missing values.

```{r fill_na, echo=TRUE}

dat_fill <- dat %>%
  merge(dat_int, by="interval", all.x=TRUE) %>%
  mutate(steps=ifelse(is.na(steps), steps_med, steps))

```

```{r histogram_updated, echo=TRUE}

dat_day_upd <- dat_fill %>%
  filter(!is.na(steps)) %>%
  group_by(date) %>%
  summarise(steps_day=sum(steps))

g <- ggplot(dat_day_upd, aes(x=date, y=steps_day))
g+geom_col()+theme_bw()+labs(title="Total number of steps per day")


mean_upd <- mean(dat_day_upd$steps_day)
median_upd <- median(dat_day$steps_day)


```
After inputing missing data with the median value, the updated mean and median are:

* The mean total number of steps taken per day is `r mean_upd` instead of `r mean`.
* The median total number of steps taken per day is `r median_upd` instead of `r median`.

The mean is lower after inputing missing values, while the median stays the same. This is obvious as I used the median to fill inn, and the median is therefore unaffected. The median value is lower than the mean, and has therefore reduced the new mean value. 




## Are there differences in activity patterns between weekdays and weekends?


```{r weekdays, echo=TRUE}

weekdays_dat <- dat_fill %>%
  mutate(wd=weekdays(date),
         wd=ifelse(wd=="lørdag", "Weekend", ifelse(wd=="søndag", "Weekend", "Weekday"))) %>%
  group_by(wd, interval) %>%
  summarise(m=mean(steps))

g_wd <- ggplot(weekdays_dat, aes(x=interval, y=m))
g_wd+geom_line()+theme_bw()+labs(title="Total number of steps per day",y="Number of steps")+facet_wrap(~wd)

```

On weekdays, activity tends to start earlier in the day, but stay on a lower level throughout the day. In the weekends activity starts later in the morning, but is on a higher level mid-day. 